Measurement of nuclear excitation functions for proton induced reactions (E\texorpdfstring{$_{\text{p}}$\,=\,}{Ep = 40--90}40--90 MeV) on natural Nb

### 2018-04-03 nb_px_paper/nb_px_paper.tex
Shortened discussion of other novel production pathways.

In addition to the $^\text{nat}$Nb(p,x)\ce{^{90}Mo} measurement, this experiment has also yielded measurements of  a number of additional  emerging radionuclides with medical applications.
These include the non-standard positron emitters \ce{^{57}Ni}, \ce{^{64}Cu},  \ce{^{86}Y}, \ce{^{89}Zr},  \ce{^{90}Nb}, 
% the $\beta^-$-therapy agent  \ce{^{64}Cu},  
and the Auger-therapy agent \ce{^{82\text{m}}Rb}.

\ce{^{57}Ni} ($t_{1/2}=35.60\pm0.06$ h, $\epsilon$=100\% to \ce{^{57}Co} \cite{Bhat1998}), while useful on its own as a positron emitter, stands poised as a particularly promising candidate for theranostic pairing with the soft $\beta^-$ emitter \ce{^{66}Ni} ($t_{1/2}=54.6\pm0.3$ h, $\beta^-$=100\% to \ce{^{66}Cu} \cite{Browne2010a}) \cite{PMID:7632762,zweit1996medium,Graves2016,Rosch2014}. 
\ce{^{nat}Cu}(p,x)\ce{^{57}Ni} would seem to be an intriguing production pathway, due to the ready availability of Cu metal as target, combined with the fact that production in this pathway strongly favors \ce{^{57}Ni} over \ce{^{56}Ni} --- indeed, the \ce{^{57}Ni}/\ce{^{56}Ni} 
%ratio of cross sections is approximately 70 at 61.58 MeV, and varies from 11-18 at the 70-90 MeV positions.
ratio of production rates is approximately 290 at 61.58 MeV, and varies from 45--75 at the 70--90 MeV positions.
The traditional route for \ce{^{57}Ni} production is via \ce{^{nat}Co}(p,3n)\ce{^{57}Ni}, but at moderate energies, suffers from more  \ce{^{56}Ni} contamination than \ce{^{nat}Cu}(p,x)  ($\sigma_\text{57Ni} / \sigma_\text{56Ni}\approx$ 10 at maximum).
Lower-energy production via \ce{^{nat}Co}(p,3n) at 24--40\,MeV is below threshold for \ce{^{56}Ni}, but has a peak cross section of approximately 10 mb, making \ce{^{nat}Cu}(p,x) the superior production route for moderate-energy accelerators  \cite{MICHEL1997153,Ditrói2013}.

% Moving away from the potential PET emitters, 
\ce{^{64}Cu}  ($t_{1/2}$ = 12.7 h) undergoes $\beta^+$ decay (61.5\% branching ratio) to \ce{^{64}Ni} or $\beta^-$ decay (38.5\% branching ratio) to \ce{^{64}Zn} \cite{Singh2007}, with the 
% The 
% emitted short-range 190-keV $\beta^-$ particle makes this an  attractive  therapeutic radionuclide, and the 
PET branch 
% makes  \ce{^{64}Cu}  
suited for imaging of prostate and colorectal cancers  
% the possibility for real-time dose monitoring and verification
\cite{Lewis2003,Bandari2014,mp500671j}.
% This makes \ce{^{64}Cu} particularly desirable  for emerging radiation therapy protocols \cite{Lewis2003,Bandari2014,mp500671j}.
Several production routes currently exist: \ce{^{64}Ni}(p,n)  uses 8--14 MeV protons on the expensive enriched target \ce{^{64}Ni} (0.9255\% natural abundance), but offers a high radioisotopic purity assuming a highly enriched target \cite{Szelecsenyi1993,Aslam2009}.
\ce{^{68}Zn}(p,$\alpha$n) requires more energetic 20--30 MeV protons, and necessitates an enriched target (18.45\% natural abundance) to avoid the co-production of radio-copper impurities \cite{Hilgers2003,Szelecsenyi2005}.
More recently, the use of compact DD neutron generators for \ce{^{nat}Zn}(n,p) production has been proposed, with the promise of mCi-scale production with high specific activity  \cite{Voyles2017}.
% \ce{^{nat}Cu}(p,x)\ce{^{64}Cu} could be another potential production pathway

\ce{^{86}Y} ($t_{1/2}=14.74\pm0.02$ h, $\epsilon$=100\% to \ce{^{86}Sr}  \cite{NEGRET20151}) is another novel  emerging  PET isotope, whose longer half-life has poised it for applications as a tracer for slower metabolic processes, as well as in   pharmacokinetics studies \cite{Valdovinos2017,Nickles2003,Qaim2008,QaimSyedM2011}.
In particular, it is highly desired to form a theranostic pair with the widely-employed $\beta^-$ therapy agent \ce{^{90}Y} ($t_{1/2}=64.00\pm0.21$ h, $\beta^-$=100\% to \ce{^{90}Zr} \cite{Browne1997}), which can be produced from a long-lived \ce{^{90}Sr} generator and emits no discrete observable gamma-rays or x-rays though decay \cite{Herzog1993}.
Although a weak positron branch exists and bremsstrahlung scintigraphy is commonly used for clinical imaging of the \ce{^{90}Y} biodistribution, theranostics  necessitate an imaging isotope to be paired for quantification of its uptake and biodistribution  \cite{Nickles2004}.
Conventional production of \ce{^{86}Y} proceeds through low-energy (7--14 MeV) irradiation via \ce{^{86}Sr}(p,n), which requires an enriched  \ce{^{86}Sr} target (9.86\% natural abundance), in order to eliminate contamination from (p,n) on the other stable \ce{^{84,87,88}Sr} isotopes  \cite{Rosch1993}.
Alternatively, production at 33--43 MeV via \ce{^{88}Sr}(p,3n) has been proposed --- this pathway also requires an enriched target (82.58\% natural abundance) for the same reason, but contamination with other Y co-activities will be even more pronounced than via (p,n), due to the opening of (p,n) and (p,2n) channels on all stable Sr isotopes \cite{doi:10.1139/v67-193,levkovski1991cross}.
Minimizing activity from other isotopes of the element in question is essential for producing radionuclides in high specific activity, as these competing isotopes are often impractical to separate out by radiochemical means.
% \comment{Stephen:  impossible by radiochemical means, and impossible by affordable/practical means.\\Substitute for ``difficult'' post-discussion.}
As a result, it would appear that \ce{^{nat}Nb}(p,x) is a poor route for  \ce{^{86}Y} production in this respect, as it only reaches a maximum of approximately 35\% radioisotopic purity.
The  dominant yttrium radioisotope produced by  \ce{^{nat}Nb}(p,x) in the 40--90 MeV region is  \ce{^{87}Y} ($t_{1/2}=79.8\pm0.3$ h, $\epsilon^-$=100\% to \ce{^{87m}Sr} \cite{Johnson2015}).
However,  \ce{^{87}Y} itself has application as a generator for  \ce{^{87m}Sr} ($t_{1/2}=2.815\pm0.012$ h, IT=99.70\% to \ce{^{87}Sr} \cite{Johnson2015}), which is used for imaging studies of metastatic bone cancers, especially when in a theranostic pair with the established therapy agent \ce{^{89}Sr} ($t_{1/2}=50.563\pm0.0025$ d, $\beta^-$=100\% to \ce{^{89}Y} \cite{Singh2013}) \cite{Kiselev1974,Kandil2009}.
% Since the radio-yttrium purity of \ce{^{87}Y} is approximately 88\% between 51--61 MeV in \ce{^{nat}Nb}(p,x), this could present an intriguing route for  \ce{^{87}Y} production.

\ce{^{89}Zr} ($t_{1/2}=78.41\pm0.12$ h, $\epsilon$=100\% to \ce{^{89}Y}  \cite{Singh2013}) is a long-lived positron emitter useful as a tracer for slow biological processes, immune studies, and imaging of liver and  breast cancers \cite{Verel2003,Dijkers2009,Dijkers2010}.
Current production focuses on \ce{^{89}Y}(p,n)\ce{^{89}Zr} between 9--14 MeV, which offers an extremely high-purity route on a mono-isotopic target and a strong population of \ce{^{89}Zr}, with a peak cross section of nearly 800 mb   \cite{PhysRevC.38.1624,Omara2009}.
Due to co-production of additional \ce{^{86,87,88}Zr} radio-zirconium,  \ce{^{nat}Nb}(p,x) is clearly inferior to  \ce{^{89}Y}(p,n)\ce{^{89}Zr}, as the Nb route has a smaller peak cross section of approximately 290 mb, and achieves only 10--20\% radioisotopic purity in the 50--90 MeV region.

\ce{^{90}Nb} ($t_{1/2}=14.60 \pm 0.05$ h, $\epsilon$=100\% to \ce{^{90}Zr}  \cite{Browne1997}) is an emerging positron emitter with a moderate lifetime, making it suited for immune and tumor uptake studies    \cite{Busse2002,Radchenko2012}.
It is typically produced using 8--15 MeV protons via \ce{^{90}Zr}(p,n)\ce{^{90}Nb}, using an enriched target (51.45\% natural abundance) for high radioisotopic purity, and produces a product with minimal contamination and a peak cross section of approximately 750 mb  \cite{Busse2002}.
\ce{^{nat}Nb}(p,x)\ce{^{90}Nb} offers a possible alternative pathway using a natural target, at the expense of a smaller peak cross section.
\ce{^{90}Nb} may be produced directly with an approximately 370 mb peak cross section and 99\% radioisotopic purity, or could be produced as a \ce{^{90}Mo}/\ce{^{90}Nb} generator, which would have nearly 100\% radioisotopic purity by using protons below the \ce{^{nat}Nb}(p,5n) threshold of 45.76 MeV.
However, the greatest problem with using the \ce{^{nat}Nb}(p,x) reaction to produce \ce{^{90}Nb} is the inability to separate the radioisotope from the target itself, rendering the production of a high-specific activity product impossible.

Finally, \ce{^{82\text{m}}Rb} ($t_{1/2}=6.472\pm0.006$ h, $\epsilon$=100\% to \ce{^{82}Kr}  \cite{Tuli2003}) is a diagnostic and emerging Auger-therapy agent, typically seen as a contaminant in \ce{^{82}Sr}/\ce{^{82}Rb} generators 
% for PET studies
\cite{Kovacs1991}.
It is commonly produced via \ce{^{82}Kr}(p,n) at 10--15 MeV, using an enriched \ce{^{82}Kr} gaseous target, with a peak cross section of approximately 400 mb at 12 MeV \cite{Kovacs1991}.
Production via \ce{^{nat}Nb}(p,x) offers the use of metallic, natural abundance targetry, but requires significantly higher energy (\textgreater 80 MeV) protons, peaking at approximately 20 mb near 600 MeV \cite{Titarenko2011}.
It is clear that this production route offers no advantage over existing \ce{^{82}Kr}(p,n) routes for in-house production.

>>>>

In addition to the $^\text{nat}$Nb(p,x)\ce{^{90}Mo} measurement, this experiment has also yielded measurements of  a number of additional  emerging radionuclides with medical applications.
These include the non-standard positron emitters 
\ce{^{57}Ni} \cite{PMID:7632762,zweit1996medium,Graves2016,Rosch2014}, 
\ce{^{64}Cu} \cite{Lewis2003,Bandari2014,mp500671j,Szelecsenyi1993,Aslam2009,Hilgers2003,Szelecsenyi2005,Voyles2017},  \ce{^{86}Y} \cite{Valdovinos2017,Nickles2003,Qaim2008,QaimSyedM2011,Rosch1993,doi:10.1139/v67-193,levkovski1991cross,Johnson2015,Singh2013,Kiselev1974,Kandil2009}, 
\ce{^{89}Zr}  \cite{Verel2003,Dijkers2009,Dijkers2010,PhysRevC.38.1624,Omara2009},  
\ce{^{90}Nb} \cite{Busse2002,Radchenko2012},  
% the $\beta^-$-therapy agent  \ce{^{64}Cu},  
and the Auger-therapy agent \ce{^{82\text{m}}Rb} \cite{Kovacs1991,Titarenko2011}. 
Production of these radionuclides offers no major advantages over established pathways, with the generally lower yields and radioisotopic purities failing to justify the convenience of natural targets  via   \ce{^{nat}Cu}(p,x) and  \ce{^{nat}Nb}(p,x). 
The one possible exception to this trend is the non-standard positron emitter \ce{^{57}Ni} ($t_{1/2}=35.60\pm0.06$ h, $\epsilon$=100\% to \ce{^{57}Co} \cite{Bhat1998}) --- the \ce{^{57}Ni}/\ce{^{56}Ni} ratio of production rates is approximately 290 at 61.58 MeV, and varies from 45--75 at the 70--90 MeV positions.
This \ce{^{nat}Cu}(p,x) route offers both higher yield and higher radioisotopic purity over the established  \ce{^{nat}Co}(p,3n) pathway, which suffers from approximately five-fold greater  \ce{^{56}Ni} contamination \cite{MICHEL1997153,Ditrói2013}.

% %%%
% %
% %  Moving this section to PhD thesis - too much detail on applications for an experimental paper
% %
% %%%
% 
% \ce{^{57}Ni} ($t_{1/2}=35.60\pm0.06$ h, $\epsilon$=100\% to \ce{^{57}Co} \cite{Bhat1998}), while useful on its own as a positron emitter, stands poised as a particularly promising candidate for theranostic pairing with the soft $\beta^-$ emitter \ce{^{66}Ni} ($t_{1/2}=54.6\pm0.3$ h, $\beta^-$=100\% to \ce{^{66}Cu} \cite{Browne2010a}) \cite{PMID:7632762,zweit1996medium,Graves2016,Rosch2014}. 
% \ce{^{nat}Cu}(p,x)\ce{^{57}Ni} would seem to be an intriguing production pathway, due to the ready availability of Cu metal as target, combined with the fact that production in this pathway strongly favors \ce{^{57}Ni} over \ce{^{56}Ni} --- indeed, the \ce{^{57}Ni}/\ce{^{56}Ni} 
% %ratio of cross sections is approximately 70 at 61.58 MeV, and varies from 11-18 at the 70-90 MeV positions.
% ratio of production rates is approximately 290 at 61.58 MeV, and varies from 45--75 at the 70--90 MeV positions.
% The traditional route for \ce{^{57}Ni} production is via \ce{^{nat}Co}(p,3n)\ce{^{57}Ni}, but at moderate energies, suffers from more  \ce{^{56}Ni} contamination than \ce{^{nat}Cu}(p,x)  ($\sigma_\text{57Ni} / \sigma_\text{56Ni}\approx$ 10 at maximum).
% Lower-energy production via \ce{^{nat}Co}(p,3n) at 24--40\,MeV is below threshold for \ce{^{56}Ni}, but has a peak cross section of approximately 10 mb, making \ce{^{nat}Cu}(p,x) the superior production route for moderate-energy accelerators  \cite{MICHEL1997153,Ditrói2013}.
% 
% 
% % Moving away from the potential PET emitters, 
% \ce{^{64}Cu}  ($t_{1/2}$ = 12.7 h) undergoes $\beta^+$ decay (61.5\% branching ratio) to \ce{^{64}Ni} or $\beta^-$ decay (38.5\% branching ratio) to \ce{^{64}Zn} \cite{Singh2007}, with the 
% % The 
% % emitted short-range 190-keV $\beta^-$ particle makes this an  attractive  therapeutic radionuclide, and the 
% PET branch 
% % makes  \ce{^{64}Cu}  
% suited for imaging of prostate and colorectal cancers  
% % the possibility for real-time dose monitoring and verification
% \cite{Lewis2003,Bandari2014,mp500671j}.
% % This makes \ce{^{64}Cu} particularly desirable  for emerging radiation therapy protocols \cite{Lewis2003,Bandari2014,mp500671j}.
% Several production routes currently exist: \ce{^{64}Ni}(p,n)  uses 8--14 MeV protons on the expensive enriched target \ce{^{64}Ni} (0.9255\% natural abundance), but offers a high radioisotopic purity assuming a highly enriched target \cite{Szelecsenyi1993,Aslam2009}.
% \ce{^{68}Zn}(p,$\alpha$n) requires more energetic 20--30 MeV protons, and necessitates an enriched target (18.45\% natural abundance) to avoid the co-production of radio-copper impurities \cite{Hilgers2003,Szelecsenyi2005}.
% More recently, the use of compact DD neutron generators for \ce{^{nat}Zn}(n,p) production has been proposed, with the promise of mCi-scale production with high specific activity  \cite{Voyles2017}.
% % \ce{^{nat}Cu}(p,x)\ce{^{64}Cu} could be another potential production pathway 
% 
% 
% 
% \ce{^{86}Y} ($t_{1/2}=14.74\pm0.02$ h, $\epsilon$=100\% to \ce{^{86}Sr}  \cite{NEGRET20151}) is another novel  emerging  PET isotope, whose longer half-life has poised it for applications as a tracer for slower metabolic processes, as well as in   pharmacokinetics studies \cite{Valdovinos2017,Nickles2003,Qaim2008,QaimSyedM2011}.
% In particular, it is highly desired to form a theranostic pair with the widely-employed $\beta^-$ therapy agent \ce{^{90}Y} ($t_{1/2}=64.00\pm0.21$ h, $\beta^-$=100\% to \ce{^{90}Zr} \cite{Browne1997}), which can be produced from a long-lived \ce{^{90}Sr} generator and emits no discrete observable gamma-rays or x-rays though decay \cite{Herzog1993}.
% Although a weak positron branch exists and bremsstrahlung scintigraphy is commonly used for clinical imaging of the \ce{^{90}Y} biodistribution, theranostics  necessitate an imaging isotope to be paired for quantification of its uptake and biodistribution  \cite{Nickles2004}.
% Conventional production of \ce{^{86}Y} proceeds through low-energy (7--14 MeV) irradiation via \ce{^{86}Sr}(p,n), which requires an enriched  \ce{^{86}Sr} target (9.86\% natural abundance), in order to eliminate contamination from (p,n) on the other stable \ce{^{84,87,88}Sr} isotopes  \cite{Rosch1993}.
% Alternatively, production at 33--43 MeV via \ce{^{88}Sr}(p,3n) has been proposed --- this pathway also requires an enriched target (82.58\% natural abundance) for the same reason, but contamination with other Y co-activities will be even more pronounced than via (p,n), due to the opening of (p,n) and (p,2n) channels on all stable Sr isotopes \cite{doi:10.1139/v67-193,levkovski1991cross}.
% Minimizing activity from other isotopes of the element in question is essential for producing radionuclides in high specific activity, as these competing isotopes are often impractical to separate out by radiochemical means.
% % \comment{Stephen:  impossible by radiochemical means, and impossible by affordable/practical means.\\Substitute for ``difficult'' post-discussion.}
% As a result, it would appear that \ce{^{nat}Nb}(p,x) is a poor route for  \ce{^{86}Y} production in this respect, as it only reaches a maximum of approximately 35\% radioisotopic purity.
% The  dominant yttrium radioisotope produced by  \ce{^{nat}Nb}(p,x) in the 40--90 MeV region is  \ce{^{87}Y} ($t_{1/2}=79.8\pm0.3$ h, $\epsilon^-$=100\% to \ce{^{87m}Sr} \cite{Johnson2015}).
% However,  \ce{^{87}Y} itself has application as a generator for  \ce{^{87m}Sr} ($t_{1/2}=2.815\pm0.012$ h, IT=99.70\% to \ce{^{87}Sr} \cite{Johnson2015}), which is used for imaging studies of metastatic bone cancers, especially when in a theranostic pair with the established therapy agent \ce{^{89}Sr} ($t_{1/2}=50.563\pm0.0025$ d, $\beta^-$=100\% to \ce{^{89}Y} \cite{Singh2013}) \cite{Kiselev1974,Kandil2009}.
% % Since the radio-yttrium purity of \ce{^{87}Y} is approximately 88\% between 51--61 MeV in \ce{^{nat}Nb}(p,x), this could present an intriguing route for  \ce{^{87}Y} production.
% 
% 
% 
% 
% \ce{^{89}Zr} ($t_{1/2}=78.41\pm0.12$ h, $\epsilon$=100\% to \ce{^{89}Y}  \cite{Singh2013}) is a long-lived positron emitter useful as a tracer for slow biological processes, immune studies, and imaging of liver and  breast cancers \cite{Verel2003,Dijkers2009,Dijkers2010}.
% Current production focuses on \ce{^{89}Y}(p,n)\ce{^{89}Zr} between 9--14 MeV, which offers an extremely high-purity route on a mono-isotopic target and a strong population of \ce{^{89}Zr}, with a peak cross section of nearly 800 mb   \cite{PhysRevC.38.1624,Omara2009}.
% Due to co-production of additional \ce{^{86,87,88}Zr} radio-zirconium,  \ce{^{nat}Nb}(p,x) is clearly inferior to  \ce{^{89}Y}(p,n)\ce{^{89}Zr}, as the Nb route has a smaller peak cross section of approximately 290 mb, and achieves only 10--20\% radioisotopic purity in the 50--90 MeV region.
% 
% 
% 
% 
% \ce{^{90}Nb} ($t_{1/2}=14.60 \pm 0.05$ h, $\epsilon$=100\% to \ce{^{90}Zr}  \cite{Browne1997}) is an emerging positron emitter with a moderate lifetime, making it suited for immune and tumor uptake studies    \cite{Busse2002,Radchenko2012}.
% It is typically produced using 8--15 MeV protons via \ce{^{90}Zr}(p,n)\ce{^{90}Nb}, using an enriched target (51.45\% natural abundance) for high radioisotopic purity, and produces a product with minimal contamination and a peak cross section of approximately 750 mb  \cite{Busse2002}.
% \ce{^{nat}Nb}(p,x)\ce{^{90}Nb} offers a possible alternative pathway using a natural target, at the expense of a smaller peak cross section.
% \ce{^{90}Nb} may be produced directly with an approximately 370 mb peak cross section and 99\% radioisotopic purity, or could be produced as a \ce{^{90}Mo}/\ce{^{90}Nb} generator, which would have nearly 100\% radioisotopic purity by using protons below the \ce{^{nat}Nb}(p,5n) threshold of 45.76 MeV.
% However, the greatest problem with using the \ce{^{nat}Nb}(p,x) reaction to produce \ce{^{90}Nb} is the inability to separate the radioisotope from the target itself, rendering the production of a high-specific activity product impossible.  
% 
% 
% 
% 
% 
% 
% 
% Finally, \ce{^{82\text{m}}Rb} ($t_{1/2}=6.472\pm0.006$ h, $\epsilon$=100\% to \ce{^{82}Kr}  \cite{Tuli2003}) is a diagnostic and emerging Auger-therapy agent, typically seen as a contaminant in \ce{^{82}Sr}/\ce{^{82}Rb} generators 
% % for PET studies
% \cite{Kovacs1991}.
% It is commonly produced via \ce{^{82}Kr}(p,n) at 10--15 MeV, using an enriched \ce{^{82}Kr} gaseous target, with a peak cross section of approximately 400 mb at 12 MeV \cite{Kovacs1991}.
% Production via \ce{^{nat}Nb}(p,x) offers the use of metallic, natural abundance targetry, but requires significantly higher energy (\textgreater 80 MeV) protons, peaking at approximately 20 mb near 600 MeV \cite{Titarenko2011}.
% It is clear that this production route offers no advantage over existing \ce{^{82}Kr}(p,n) routes for in-house production.
